Whole-body connectome of a segmented annelid larva

### 2024-03-17 Manuscript.qmd
MS: render w.o. toc, upload suppl table 2

---
title: "Whole-body connectome of a segmented annelid larva"
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
crossref:
  fig-title: Figure     # (default is "Figure")
max-width: 100%
format:
  html: default
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---
\

@@

The EM image stack with all traces and annotations are available at <https://catmaid.jekelylab.ex.ac.uk> and can be queried in CATMAID or via the CATMAID application programming interface (API) with the catmaid/natverse packages [@bates2020natverse].
The dataset includes all EM images (in jpg format), skeletons, meshes, node tags, connectors (both synapses and desmosomes) and annotations.
We also provide all the R scripts we used for data retrieval and the generation of the figures [add Github/Zenodo repo]. All plots, diagrams (including anatomical renderings), figure layouts should be fully reproducible by using the R scripts provided. Scripts were mostly organised by figure, except some generalist scripts e.g. to load libraries, CATMAID access data and general functions ('Natverse_functions_and_conn.R').

@@

Due to contrast and focus problems in the main dataset we had to re-image certain layers at higher resolutions, to allow tracing of neurons.
These re-imaged series were made into independent CATMAID projects.
This included five extra projects, taken at various points throughout the main dataset [(Supplemental Table xy - Reza)](check).
The largest of these, Plexus_HT-4_Naomi_project\_\_372-4013, consisted of 1407 layers at resolution of 2.2 nm.
This stack mostly focused on the brain plexus and the ventral nerve cord where most neurites and synapses occur.
Other projects consisted of three jump/gap regions that required not only high resolution but also better realignment.
One set contained all the immunogold labelled layers [@shahidi2015serial] that were not included in the main aligned dataset.
These layers had very low contrast due to the immunolabelling procedure and therefore required higher resolution imaging.
All projects were first created and processed in TrakEM2 and then exported as flat jpeg images into CATMAID.

>>>>

---
title: "Whole-body connectome of a segmented annelid larva"
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
crossref:
  fig-title: Figure     # (default is "Figure")
max-width: 100%
format:
  html: default
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: false
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---
\

@@

The EM image stack with all traces and annotations are available at <https://catmaid.jekelylab.ex.ac.uk> and can be queried in CATMAID or via the CATMAID application programming interface (API) with the catmaid/natverse packages [@bates2020natverse].
The dataset includes all EM images (in jpg format), skeletons, meshes, node tags, connectors (both synapses and desmosomes) and annotations.
We also provide all the R scripts we used for data retrieval and the generation of the figures [@jekely_2024_10825371]. All plots, diagrams (including anatomical renderings), figure layouts should be fully reproducible by using the R scripts provided. Scripts were mostly organised by figure, except some generalist scripts e.g. to load libraries, CATMAID access data and general functions ('Natverse_functions_and_conn.R').

@@

Due to contrast and focus problems in the main dataset we had to re-image certain layers at higher resolutions, to allow tracing of neurons.
These re-imaged series were made into independent CATMAID projects.
This included five extra projects, taken at various points throughout the main dataset. All layers including the information on lost layers, immunogold labelling and re-imaging are listed in Supplementary Table 2.
The largest of these, Plexus_HT-4_Naomi_project\_\_372-4013, consisted of 1407 layers at resolution of 2.2 nm.
This stack mostly focused on the brain plexus and the ventral nerve cord where most neurites and synapses occur.
Other projects consisted of three jump/gap regions that required not only high resolution but also better realignment.
One set contained all the immunogold labelled layers [@shahidi2015serial] that were not included in the main aligned dataset.
These layers had very low contrast due to the immunolabelling procedure and therefore required higher resolution imaging.
All projects were first created and processed in TrakEM2 and then exported as flat jpeg images into CATMAID.

### 2024-03-17 Manuscript.tex
MS: render w.o. toc, upload suppl table 2

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}
\setstretch{1.2}
\hfill\break

@@

\textsuperscript{1}Heidelberg University, Centre for Organismal Studies
(COS), 69120 Heidelberg, Germany\\
\textsuperscript{2}Living Systems Institute, University of Exeter,
Exeter, EX4 4QD, United Kingdom\\
\textsuperscript{3}School of Biological Sciences, University of Bristol,
Bristol, United Kingdom\\
\textsuperscript{4}BioSciences, University of Exeter, Stocker Road,
Exeter EX4 4QD, Exeter, UK \textsuperscript{5}Electron Microscopy Core
Facility (EMCF), University of Heidelberg, 69120 Heidelberg, Germany
\textsuperscript{*}Correspondence: gaspar.jekely@cos.uni-heidelberg.de

@@

We previously reported synaptic connectomes for several whole-body
circuits from a transmission EM volume of a three-day-old larva (Randel
et al., 2015). These include the visual, startle, ciliomotor, nuchal
organ, and neurosecretory systems (Bezares-Calderón et al., 2018; Randel
et al., 2015; Shahidi et al., 2015; Verasztó et al., 2018, 2017;
Williams et al., 2017). Here, we report the complete synaptic connectome
and the cell-type complement from the same volume of a three-day-old
\emph{Platynereis} larva. Recently we also reported the desmosomal
connectome and motoneuron innervation of the somatic musculature (Jasek
et al., 2022). The full connectome reconstruction has now allowed us to
uncover several new circuits and to consider all circuits in a
whole-body context. We also found several neurons that span the entire
length of the larva, highlighting the strength of a whole-body dataset.
These cells and their circuits allow us to generate hypotheses on how
whole-body coordination may be achieved by the larval nervous system.
The connectome also allows us to address long-standing hypotheses about
the origin of the segmented annelid body-plan and explore patterns of
circuit evolution.

@@

To identify more strongly connected subgraphs within the connectome, we
used community detection with the Leiden algorithm (Traag et al., 2019).
This analysis combined with force-field-based clustering highlighted
several modules (Figures 2). Nodes in a module are more strongly
interconnected than to nodes in other modules (Figure 2---figure
supplement 3). We named the modules based on their primary effector
organs or other dominant anatomical characters. We would like to note
that this analysis can recover different number of modules depending on
the resolution parameter (Figure 2---figure supplement 2C). The selected
parameters results in communities that we think best represent
functional units.

@@

The modules are also interconnected among themselves suggesting
crosstalk. Most sensory neurons occur in the visual, the two
mechanosensory, the mushroom body and the anterior neurosecretory
modules (Figure 2---figure supplement 1J). In a Sankey diagram
representing information flow, these modules are upstream of the
motor-dominated pigmentmotor, ciliomotor, and musclemotor modules
(Figure 2---figure supplement 2B).

@@

To explore sensory-motor pathways in the grouped graph, we searched for
the shortest directed paths from all sensory neurons to all effector
cells. We identified four sensory neurons with direct motor output
(sensory-motor neurons: eyespot-PRCR3, pygPBunp, pygCirrusPUSM,
hPU2l-asymPDF) and 25 premotor sensory neurons. The maximum number of
hops from sensory neurons to an effector is five. 33 sensory neurons
have no synaptic paths to effectors. 19 of these are neurosecretory
cells with few or no synaptic partners (Williams et al., 2017)
(Figure4---figure supplement 5). For a graph-based visualisation of
information flow, we arranged the nodes according to their relative
ration of incoming and outgoing synapses (Figure 4---figure supplement
6). This layout indicates that sensory-evoked activity propagates from
sensory neurons through interneurons and motoneurons to effectors.

@@

Ranking trunk and pygidialmaps cell types by various network centrality
measures often identified segment-specific cells as strongly connected
(e.g.~pygPBunp, Ser-tr1, Loop)(Figure 12---figure supplement 2).

@@

The strongest postsynaptic partners of the diverse girdle mechanosensory
neurons are ipsilateral interneurons with a bifurcating projection
(presudounipolar morphology). This family of interneurons ---
collectively referred to as INsplit --- could be subdivided into several
distinct cell types (INsplitCR, INsplitPB, INsplitPBant,
INsplitPB\_RF/Ya, INsplitPUh, INsplitBronto, INCM)(Figure
16A-H)(Bezares-Calderón et al., 2018).

@@

The EM image stack with all traces and annotations are available at
\url{https://catmaid.jekelylab.ex.ac.uk} and can be queried in CATMAID
or via the CATMAID application programming interface (API) with the
catmaid/natverse packages (Bates et al., 2020). The dataset includes all
EM images (in jpg format), skeletons, meshes, node tags, connectors
(both synapses and desmosomes) and annotations. We also provide all the
R scripts we used for data retrieval and the generation of the figures
{[}add Github/Zenodo repo{]}. All plots, diagrams (including anatomical
renderings), figure layouts should be fully reproducible by using the R
scripts provided. Scripts were mostly organised by figure, except some
generalist scripts e.g.~to load libraries, CATMAID access data and
general functions (`Natverse\_functions\_and\_conn.R').

@@

Due to contrast and focus problems in the main dataset we had to
re-image certain layers at higher resolutions, to allow tracing of
neurons. These re-imaged series were made into independent CATMAID
projects. This included five extra projects, taken at various points
throughout the main dataset \href{check}{(Supplemental Table xy -
Reza)}. The largest of these, Plexus\_HT-4\_Naomi\_project\_\_372-4013,
consisted of 1407 layers at resolution of 2.2 nm. This stack mostly
focused on the brain plexus and the ventral nerve cord where most
neurites and synapses occur. Other projects consisted of three jump/gap
regions that required not only high resolution but also better
realignment. One set contained all the immunogold labelled layers
(Shahidi et al., 2015) that were not included in the main aligned
dataset. These layers had very low contrast due to the immunolabelling
procedure and therefore required higher resolution imaging. All projects
were first created and processed in TrakEM2 and then exported as flat
jpeg images into CATMAID.

@@

@@

>>>>

\setstretch{1.2}
\hfill\break

@@

\textsuperscript{1}Heidelberg University, Centre for Organismal Studies
(COS), 69120 Heidelberg, Germany\\
\textsuperscript{2}Living Systems Institute, University of Exeter,
Exeter, EX4 4QD, United Kingdom\\
\textsuperscript{3}School of Biological Sciences, University of Bristol,
Bristol, United Kingdom\\
\textsuperscript{4}BioSciences, University of Exeter, Stocker Road,
Exeter EX4 4QD, Exeter, United Kingdom\\
\textsuperscript{5}Electron Microscopy Core Facility (EMCF), University
of Heidelberg, 69120 Heidelberg, Germany\\
\textsuperscript{*}Correspondence: gaspar.jekely@cos.uni-heidelberg.de

@@

We previously reported synaptic connectomes for several whole-body
circuits from a transmission EM volume of a three-day-old larva (Randel
et al., 2015). These include the visual, startle, ciliomotor, nuchal
organ, and neurosecretory systems (Bezares-Calderón et al., 2018; Randel
et al., 2015; Shahidi et al., 2015; Verasztó et al., 2018, 2017;
Williams et al., 2017). Here, we report the complete synaptic connectome
and the cell-type complement from the same volume of a three-day-old
\emph{Platynereis} larva. Recently we also reported the desmosomal
connectome and motoneuron innervation of the somatic musculature (Jasek
et al., 2022) and a preprint of a preliminary analysis of the synaptic
connectome (Verasztó et al., 2020) that this work supersedes. The full
connectome reconstruction has now allowed us to uncover several new
circuits and to consider all circuits in a whole-body context. We also
found several neurons that span the entire length of the larva,
highlighting the strength of a whole-body dataset. These cells and their
circuits allow us to generate hypotheses on how whole-body coordination
may be achieved by the larval nervous system. The connectome also allows
us to address long-standing hypotheses about the origin of the segmented
annelid body-plan and explore patterns of circuit evolution.

@@

To identify more strongly connected subgraphs within the connectome, we
used community detection with the Leiden algorithm (Traag et al., 2019).
This analysis combined with force-field-based clustering highlighted
several modules (Figures 2). Nodes in a module are more strongly
interconnected than to nodes in other modules (Figure 2---figure
supplement 3). We named the modules based on their primary effector
organs or other dominant anatomical characters. We would like to note
that this analysis can recover varying numbers of modules depending on
the resolution parameter (Figure 2---figure supplement 2C). The selected
resolution parameter subdivides the network into communities that we
think best represent functional units.

@@

The modules are also interconnected among themselves suggesting
crosstalk. Most sensory neurons occur in the visual, the two
mechanosensory, the mushroom body and the anterior neurosecretory
modules (Figure 2---figure supplement 1J). In a Sankey network diagram
representing information flow, these modules are upstream of the
motor-dominated pigmentmotor, ciliomotor, and musclemotor modules
(Figure 2---figure supplement 2B).

@@

To explore sensory-motor pathways in the grouped graph, we searched for
the shortest directed paths from all sensory neurons to all effector
cells. We identified four sensory neurons with direct motor output
(sensory-motor neurons: eyespot-PRCR3, pygPBunp, pygCirrusPUSM,
hPU2l-asymPDF) and 25 premotor sensory neurons. The maximum number of
hops from sensory neurons to an effector is five. 33 sensory neurons
have no synaptic paths to effectors. 19 of these are neurosecretory
cells with few or no synaptic partners (Williams et al., 2017) (Figure
4---figure supplement 5). For a graph-based visualisation of information
flow, we arranged the nodes according to their relative ratio of
incoming and outgoing synapses (Figure 4---figure supplement 6). This
layout indicates that sensory-evoked activity propagates from sensory
neurons through interneurons and motoneurons to effectors.

@@

Ranking trunk and pygidial cell types by various network centrality
measures often identified segment-specific cells as strongly connected
(e.g.~pygPBunp, Ser-tr1, Loop)(Figure 12---figure supplement 2).

@@

The strongest postsynaptic partners of the diverse girdle mechanosensory
neurons are ipsilateral interneurons with a bifurcating projection
(presudounipolar morphology). This family of interneurons ---
collectively referred to as INsplit --- could be subdivided into several
distinct cell types (INsplitCR, INsplitPB, INsplitPBant,
INsplitPB-RF/Ya, INsplitPUh, INsplitBronto, INCM)(Figure
16A-H)(Bezares-Calderón et al., 2018).

@@

The EM image stack with all traces and annotations are available at
\url{https://catmaid.jekelylab.ex.ac.uk} and can be queried in CATMAID
or via the CATMAID application programming interface (API) with the
catmaid/natverse packages (Bates et al., 2020). The dataset includes all
EM images (in jpg format), skeletons, meshes, node tags, connectors
(both synapses and desmosomes) and annotations. We also provide all the
R scripts we used for data retrieval and the generation of the figures
(Jékely et al., 2024). All plots, diagrams (including anatomical
renderings), figure layouts should be fully reproducible by using the R
scripts provided. Scripts were mostly organised by figure, except some
generalist scripts e.g.~to load libraries, CATMAID access data and
general functions (`Natverse\_functions\_and\_conn.R').

@@

Due to contrast and focus problems in the main dataset we had to
re-image certain layers at higher resolutions, to allow tracing of
neurons. These re-imaged series were made into independent CATMAID
projects. This included five extra projects, taken at various points
throughout the main dataset. All layers including the information on
lost layers, immunogold labelling and re-imaging are listed in
Supplementary Table 2. The largest of these,
Plexus\_HT-4\_Naomi\_project\_\_372-4013, consisted of 1407 layers at
resolution of 2.2 nm. This stack mostly focused on the brain plexus and
the ventral nerve cord where most neurites and synapses occur. Other
projects consisted of three jump/gap regions that required not only high
resolution but also better realignment. One set contained all the
immunogold labelled layers (Shahidi et al., 2015) that were not included
in the main aligned dataset. These layers had very low contrast due to
the immunolabelling procedure and therefore required higher resolution
imaging. All projects were first created and processed in TrakEM2 and
then exported as flat jpeg images into CATMAID.

@@

\bibitem[\citeproctext]{ref-jekely_2024_10825371}
Jékely G, Jasek S, Gühmann M, Bezares-Calderón LA, Williams EA, Shahidi
R. 2024. {Code documentation for the Jekely et al. three- day-old
Platynereis larva connectome paper}.
doi:\href{https://doi.org/10.5281/zenodo.10825371}{10.5281/zenodo.10825371}

@@

\bibitem[\citeproctext]{ref-veraszto2020}
Verasztó C, Jasek S, Gühmann M, Shahidi R, Ueda N, Beard JD, Mendes S,
Heinz K, Bezares-Calderón LA, Williams E, Jékely G. 2020.
\href{http://dx.doi.org/10.1101/2020.08.21.260984}{Whole-animal
connectome and cell-type complement of the three-segmented
{\emph{Platynereis dumerilii}} larva}.
